---
title: "R-Adventure using R-Notebook AND Iris dataset"
output:
  pdf_document:
    highlight: tango
    toc: yes
  html_notebook:
    toc: yes
---

# Load library

```{r}
library(caret)
```

# Load data

Influenced by \| based on: [R Machine Learning](https://machinelearningmastery.com/machine-learning-in-r-step-by-step/) By **Jason Brownlee** published on *February 3, 2016*. [other](https://www.r-bloggers.com/2022/02/beginners-guide-to-machine-learning-in-r-with-step-by-step-tutorial/)

```{r}
# attach the iris dataset to the environment
data(iris)
# rename the dataset
dataset <- iris

print(dataset)
```

# Preprocessing

## train-test split

```{r}
# create a list of 80% of the rows in the original dataset we can use for training
validation_index <- createDataPartition(dataset$Species, p=0.80, list=FALSE)
# select 20% of the data for validation
validation <- dataset[-validation_index,]
# use the remaining 80% of data to training and testing the models
dataset <- dataset[validation_index,]
```

# EDA

## stats

```{r}
# dimensions of dataset
print("data dimension")
dim(dataset)
# list types for each attribute
print("data types")
sapply(dataset, class)
```

```{r}
# take a peek at the first 5 rows of the data
print(head(dataset))
```

```{r}
# list the levels for the class
levels(dataset$Species)
print("--------")
# summarize the class distribution
percentage <- prop.table(table(dataset$Species)) * 100
cbind(freq=table(dataset$Species), percentage=percentage)

# summarize attribute distributions
print("--------")
summary(dataset)
```

## visuals

```{r}
# split input and output
x <- dataset[,1:4]
y <- dataset[,5]

# boxplot for each attribute on one image
par(mfrow=c(1,4))
  for(i in 1:4) {
  boxplot(x[,i], sub=names(iris)[i])
  }
title(main = "boxplot for each attribute", line=-2, outer = TRUE)
```

```{r}
# barplot for class breakdown
plot(y, main="barplot for class breakdown")
```

```{r}
# scatterplot matrix
featurePlot(x=x, y=y, plot="ellipse")
```

```{r}
# box and whisker plots for each attribute
featurePlot(x=x, y=y, plot="box")
```

```{r}
# density plots for each attribute by class value
scales <- list(x=list(relation="free"), y=list(relation="free"))
featurePlot(x=x, y=y, plot="density", scales=scales)
```

# ML

## CV

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="cv", number=10)
metric <- "Accuracy"
```

## ML Models

-   Linear Discriminant Analysis (LDA)
-   Classification and Regression Trees (CART).
-   k-Nearest Neighbors (kNN).
-   Support Vector Machines (SVM) with a linear kernel.
-   Random Forest (RF)

```{r}
# a) linear algorithms
set.seed(7)
fit.lda <- train(Species~., data=dataset, method="lda", metric=metric, trControl=control)
# b) nonlinear algorithms
# CART
set.seed(7)
fit.cart <- train(Species~., data=dataset, method="rpart", metric=metric, trControl=control)
# kNN
set.seed(7)
fit.knn <- train(Species~., data=dataset, method="knn", metric=metric, trControl=control)
# c) advanced algorithms
# SVM
set.seed(7)
fit.svm <- train(Species~., data=dataset, method="svmRadial", metric=metric, trControl=control)
# Random Forest
set.seed(7)
fit.rf <- train(Species~., data=dataset, method="rf", metric=metric, trControl=control)
```

## model selection

```{r}
# summarize accuracy of models
results <- resamples(list(lda=fit.lda, cart=fit.cart, knn=fit.knn, svm=fit.svm, rf=fit.rf))
summary(results)
```

```{r}
# compare accuracy of models
dotplot(results)
```

```{r}
# summarize Best Model
print(fit.lda)
```

## prediction on test \| validation data

```{r}
# estimate skill of LDA on the validation dataset
predictions <- predict(fit.lda, validation)
confusionMatrix(predictions, validation$Species)
```
